### 2019-09-09 manuscript/manuscript.tex
ms edits

Since the seminal theoretic model of \textcite{Maynard_Smith1974-lc}
demonstrating that diversity is reduced at linked neutral variation as a
advantageous polymorphism sweeps fixation, over four decades of theoretical and
empirical research has enriched our understanding of linked selection. This
work has led to the discovery of selected genes from genomic signals of
hitchhiking \parencite{Nair2003-tw,Voight2006-rn}, found genome-wide signals of
linked selection
\parencite{Aguade1989-jx,Begun1992-ey,Cutter2010-gi,Andersen2012-bj,Cutter2003-tl}
leading to the development of recurrent hitchhiking \parencite{Stephan1992-jc}
and background selection \parencite{Charlesworth1993-gb} models that attempt
explain this pattern. Since, much attention has focused on quantifying the
impact of background selection \parencite{McVicker2009-ax} and developing
theoretic machinery \parencite{Coop2012-cd} to differentiate the regional
reductions in diversity into those caused by full and partial sweeps, and those
caused by background selection \parencite{Elyashiv2016-vt}. However, as other
theoretic models of soft sweeps from standing variation emerged
\parencite{Hermisson2005-hs,Pennings2006-lj}, attention shifted towards
detecting other modes of linked selection \parencite{Pritchard2010-tk}, a more
difficult endeavor \parencite{Przeworski2005-bg} met with new coalescent models
\parencite{Berg2015-xj} and machine learning methods
\parencite{Schrider2017-yx}. An undervalued theoretic approach to understanding
the genome-wide effects of selection on standing variation, e.g. selection on
an infinitesimal polygenic trait, stems from an early quantitative genetic
model of linked selection \parencite{Robertson1961-ho} and its later
developments \parencite{Santiago1995-hx,Santiago1998-bs}. Implicit in these
models is that autocovariance between allele frequency change is created when
there is heritable fitness variation in the population, a signal that can be
readily detected from temporal genomic data \parencite{Buffalo2019-io}. Here,
we provide the first empirical evidence of linked selection acting on standing
variation over very short timescales.

We reanalyzed three published temporal genomic datasets and detected the
genome-wide signature of linked selection acting over tens of generations.
Furthermore, we find the dynamics of temporal autocovariance are consistent
with theory; the temporal covariance in allele frequencies decays \vb{as both
  the fitness variation associated with particular haplotypes and the linkage
disequilibria between selected and neutral loci decay through time}. In
reanalyzing one study \parencite{Barghi2019-qy}, we find that after sixty
generations, greater than 20\% of the variation in allele frequency change is
directly due to the action linked selection. Capitalizing on replicated
evolve-and-resequence study designs, we characterized the extent to which
convergent selection pressures lead to parallel changes in allele frequency
across replicate populations, finding correlation in allele frequency
changes varies from 0.1 (in the \cite{Barghi2019-qy} study) to 0.36 (in the
\cite{Kelly2019-dc} study). \graham{Are these just repeat numbers from main
text? If so nix this sentence}

@@

It is worth building some intuition why temporal covariance allows us to detect
such faint signals of polygenic linked selection from temporal genomic data.
Each variant is subject to both variance in allele frequency due to drift and
two levels of sampling noise (as individuals are sampled from the population
and reads are sampled from amplified DNA), which at any locus swamps the
temporal covariance signal and creates spurious covariances. However, these
spurious covariances do not share a common sign across timepoints as the
covariances created by linked selection do; consequently, averaging across the
entire genome, the temporal signal exceeds sampling noise.

One limitation of our current work here is that none of the studies we
reanalyzed had linkage disequilibria data for the evolved populations. Our
theory of temporal autocovariance tells us that the temporal autocovariances a
neutral site experiences is determined by the product of the expected linkage
disequilibrium and additive fitness variation that persists through the
generations \parencite{Buffalo2019-ion}. This leads to a clear prediction:
regions of higher linkage disequilibrium and lower recombination should have
greater temporal autocovariance than regions with lower LD and higher
recombination. Unfortunately, we lack a high-resolution recombination map for
\emph{D. simulans} and while there are LD data for \emph{D. simulans}
\parencite{Signor2018-wg} we did not find a relationship between temporal
covariance and LD. \graham{Is important point. We'll likely need plots for
supp}

@@

Thus far, the most comprehensive studies implicating linked selection in
affecting genome-wide diversity have been in \emph{Drosophila}
\parencite{Begun1992-ey,Elyashiv2016-vt,Sattath2011-dr} and
\emph{Caenorhabditis} \cite{Cutter2003-tl,Cutter2003-tl,Andersen2012-bj},
whereas the evidence for linked selection in plant and yeast species is weak
\parencite{Cutter2013-ba}. Both \emph{Drosophila} and \emph{Caenorhabditi} have
short genetic maps (or in the case of selfing \emph{C.  elegans}, small
effective recombination rates), and large effect population sizes, two factors
predicted to increase the degree to which hitchhiking impacts genome-wide
diversity \parencite{Barton2000-zg}. However, our results here show that even
with small effective population sizes (300, 450, and 45 for the
\textcite{Kelly2019-dc}, and \textcite{Castro2019-uk} studies respectively)
around and, in the case of mice, moderately-sized genetic maps (around 14
Morgans; \cite{Cox2009-hf}), we still see a fairly strong effect of linked
selection. This suggests that while these factors may govern the dynamics of
classic hitchhiking events, they could perhaps play less of a role in polygenic
linked selection. However, with only three cases here, further studies on
different taxa are needed.

@@

Following \textcite{Waples1989-sj}, we have that that the variance in the
initial generation, which is entirely due to the binomial sampling process, is
$\var(p_0) = \nicefrac{p_0(1-p_0)}{d_0}$ where $d_0$ is the number of binomial
draws (e.g. read depth). At a later timepoint, the variance in allele frequency
is a result of both the binomial sampling process at time $t$ and the
evolutionary process.

Using the law of total variation,

\begin{align}
  \var(\widetilde{p_t}) &= \E(\var(\widetilde{p_t} | p_t)) + \var(\E(\widetilde{p_t}|p_t)) \\
                        &= \underbrace{\frac{p_t(1-p_t)}{d_t}}_\text{generation $t$ sampling noise} + \underbrace{\var(p_t)}_\text{variance due to evolutionary process}
  %\frac{\var(\widetilde{p_t})}{p_0(1-p_0)} &= \frac{p_t(1-p_t)}{p_0(1-p_0) d_t} + 1 - \left( 1-\frac{1}{2N}\right)^t.
  %\frac{\var(\widetilde{p_t})}{p_0(1-p_0)} &= \frac{p_t(1-p_t)}{p_0(1-p_0) d_t} + \frac{t}{2N} + O\left((2 N)^{-2}\right).
\end{align}

@@

\begin{align}
  \sum_{i=0}^{t-1} \var(\Delta_{_D} p_i) = \sum_{i=0}^{t-1} \frac{p_i(1-p_i)}{2N}
\end{align}

replacing the heterozygosity in generation $i$ with its expectation, we have

\begin{align}
  \sum_{i=0}^{t-1} \var(\Delta_{_D} p_i) &= p_0(1-p_0) \sum_{i=0}^{t-1} \frac{1}{2N} \left(1-\frac{1}{2N}\right)^i \\
                                         &= p_0(1-p_0) \left[1 - \left(1-\frac{1}{2N}\right)^t \right]
\end{align}

which is the usual variance in allele frequency change due to drift.  Then, the
total allele frequency change from generations $0$ to $t$ is
$\var(\widetilde{p}_t - \widetilde{p}_0) = \var(\widetilde{p}_t) +
\var(\widetilde{p}_0) - 2 \cov(\widetilde{p}_t, \widetilde{p}_0)$, where the
covariance depends on the nature of the sampling plan (see \cite{Nei1981-oy,
Waples1989-sj}). In the case where there is heritable variation for fitness,
and using the fact that $\cov(\widetilde{p}_t, \widetilde{p}_0) =
\nicefrac{p_0(1-p_0)}{2N}$ for Plan I sampling procedures
\parencite{Waples1989-sj}, we write,

\begin{align}
  \var(\widetilde{p}_t - \widetilde{p}_0) &= \var(\widetilde{p}_t) + \var(\widetilde{p}_0) - 2 C \cov(\widetilde{p}_t, \widetilde{p}_0) \\
                                          &= \frac{p_t(1-p_t)}{d_t}  + \frac{p_0(1-p_0)}{d_0} + p_0(1-p_0) \left[1 - \left(1-\frac{1}{2N}\right)^t \right] + \\ & \;\;\;\;\;\;
                                               \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i)  + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j) - \frac{C p_0(1-p_0)}{2N} \\
  \frac{\var(\widetilde{p}_t - \widetilde{p}_0)}{p_0(1-p_0)} &= 1 + \frac{p_t(1-p_t)}{p_0(1-p_0)d_t}  + \frac{1}{d_0} - \left(1-\frac{1}{2N}\right)^t + \\ & \;\;\;\;\;\;
  \sum_{i=0}^{t-1} \frac{\var(\Delta_{_H} p_i)}{p_0(1-p_0)}  + \sum_{0 \le i < j}^{t-1} \frac{\cov(\Delta p_i, \Delta p_j)}{p_0(1-p_0)} - \frac{C}{N}
\end{align}

where $C = 1$ if Plan I is used, and $C=0$ if Plan II is used (see
\cite{Waples1989-sj}, p. 380 and Figure 1 for a description of these sampling
procedures). We move terms creating a corrected estimator for the population
variance in allele frequency change, and replace all population heterozygosity
terms with the unbiased sample estimators, e.g. $\frac{d_t}{d_t-1}
\widetilde{p}_t (1- \widetilde{p}_t)$,

@@

\begin{align}
  \var(\widetilde{p}_t - \widetilde{p}_0) &= 
  p_t(1-p_t) \left(\frac{1}{n_t} + \frac{1}{d_t} - \frac{1}{n_t d_t} \right)  
  + p_0(1-p_0) \left( \frac{1}{n_0} + \frac{1}{d_0} - \frac{1}{n_0 d_0}\right)  \\ & \;\;\;\;\;\;
  - \frac{C p_0(1-p_0)}{N} + p_0(1-p_0) \left[1 - \left(1-\frac{1}{2N}\right)^t \right]+ \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i)  \\ & \;\;\;\;\;\; + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j) 
\end{align}

Through the law of total expectation, one can find that an unbiased estimator
of the heterozygosity is

\begin{align}
  \frac{n_t d_t}{(n_t-1) (d_t-1)} \widetilde{p_t}(1-\widetilde{p_t}).
\end{align}

Replacing this unbiased heterozygosity into our expression above, the total
sample variance is

\begin{align}
  \var(\widetilde{p}_t - \widetilde{p}_0) &= 
  \frac{n_t d_t \widetilde{p}_t(1-\widetilde{p}_t)}{(n_t-1)(d_t-1)} \left(\frac{1}{n_t} + \frac{1}{d_t} - \frac{1}{n_t d_t} \right) + 
 \frac{n_0 d_0 \widetilde{p}_0(1-\widetilde{p}_0)}{(n_0-1)(d_0-1)} \left( \frac{1}{n_0} + \frac{1}{d_0} - \frac{1}{n_0 d_0}\right) + \\ & \nonumber\;\;\;\;\;\;
 \frac{n_0 d_0 \widetilde{p}_0(1-\widetilde{p}_0)}{(n_0-1)(d_0-1)}   \left[1 - \left(1-\frac{1}{2N}\right)^t \right]  - \frac{C}{N}  \frac{n_0 d_0 \widetilde{p}_0(1-\widetilde{p}_0)}{(n_0-1)(d_0-1)} + \\ \nonumber & \;\;\;\;\;\; \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i)  + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j).  \\
                                                                                                                      % &= \widetilde{p}_t(1-\widetilde{p}_t)\frac{d_t + n_t - 1}{(n_t-1)(d_t-1)} + 
 % \widetilde{p}_0(1-\widetilde{p}_0)\frac{d_0 + n_0 - 1}{(n_0-1)(d_0-1)} + \\ & \nonumber\;\;\;\;\;\;
 % \widetilde{p}_0(1-\widetilde{p}_0) \frac{n_0 d_0}{(n_0-1)(d_0-1)}  \left[1 - \left(1-\frac{1}{2N}\right)^t \right] - \frac{C}{N} \widetilde{p}_0(1-\widetilde{p}_0)\frac{n_0 d_0}{(n_0-1)(d_0-1)} 
 % \\ \nonumber & \;\;\;\;\;\; + \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i)  + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j). 
\end{align}

As with equation \eqref{supp:eqn-depth-only-correction}, we can rearrange this
to get a biased-correct estimate of the variance in allele frequency change
between adjacent generations, $\var(\Delta p_t)$.

@@

We also need to apply a bias correction to the temporal covariances (and
possibly the replicate covariances if the initial sample frequencies are all
shared).

The basic issue is that $\cov(\Delta \widetilde{p}_t, \Delta
\widetilde{p}_{t+1}) = \cov(\widetilde{p}_{t+1} - \widetilde{p}_t,
\widetilde{p}_{t+2} - \widetilde{p}_{t+1})$, and thus shares the sampling noise
of timepoint $t+1$. Thus acts to bias the covariance by subtracting off the
noise variance term of $\var(\widetilde{p}_{t+1})$, so we add that back in.

\subsection{Variance-Covariance Matrix Correction}
\label{supp:matrix-correction}

Now, we extend the bias corrections for single locus variance and covariance
described in Supplementary Material Sections \ref{supp:depth-var-corr},
\ref{supp:ind-depth-var-corr}, and \ref{supp:cov-corr} to multiple sampled
loci. With frequency collected at $T+1$ timepoints across $R$ replicate
populations at $L$ loci, we have $\mathbf{F}$ of allele frequencies,
$\mathbf{D}$ multidimensional array of sequencing depths, and a $\mathbf{N}$
multidimensional array of the number of individuals sequenced, each of
dimension $R \times (T+1) \times L$.  We calculate the array $\mathbf{\Delta
F}$ which contains the allele frequency changes between adjacent generations,
and has dimension $R \times T \times L$.  The operation
$\flt(\mathbf{\Delta}\mathbf{F})$ flattens this array to a $(R \cdot T) \times
L$ matrix, such that rows are grouped by replicate, e.g. for timepoint $t$,
replicate $r$, and locus $l$ and allele frequencies $p_{t, r}$, for a single
locus the entries are

@@

\begin{align}
    \flt(\mathbf{\Delta F}) &=
                    &\begin{bmatrix} 
    \Delta p_{1, 0, 0} & \Delta p_{2, 0, 0} & \ldots & \Delta p_{1, 1, 0} & \Delta p_{2, 1, 0} & \ldots & \Delta p_{T, R, 0}  \\
    \Delta p_{1, 0, 1} & \Delta p_{2, 0, 1} & \ldots & \Delta p_{1, 1, 1} & \Delta p_{2, 1, 1} & \ldots & \Delta p_{T, R, 1}  \\
    \vdots & \vdots & \ddots & \vdots & \vdots & \ddots & \vdots  \\
    \Delta p_{1, 0, L} & \Delta p_{2, 0, L} & \ldots & \Delta p_{1, 1, L} & \Delta p_{2, 1, L} & \ldots & \Delta p_{T, R, L}  \\
  \end{bmatrix} 
\end{align}

where each $\Delta p_{t, r, l} = p_{t+1, r, l} - p_{t, r, l}$. Then, the sample
temporal-replicate covariance matrix $\mathbf{Q}'$ calculated on
$\flt(\mathbf{\Delta F})$ is a $(R \cdot T) \times (R \cdot T)$ matrix, with
the $R$ temporal-covariance block submatrices along the diagonal, and the
$R(R-1)$ replicate-covariance submatrices matrices in the upper and lower
triangles of the matrix,

\begin{align}
	\mathbf{Q}' &= 
  \begin{bmatrix} 
		\mathbf{Q}_{1,1}' & \mathbf{Q}_{1, 2}' & \ldots & \mathbf{Q}_{1, R}' \\ 
		\mathbf{Q}_{2,1}' & \mathbf{Q}_{2, 2}' & \ldots & \mathbf{Q}_{2, R}' \\ 
		\vdots & \vdots & \ddots & \vdots \\
		\mathbf{Q}_{R,1}' & \mathbf{Q}_{R, 2}' & \ldots & \mathbf{Q}_{R, R}' \\ 
  \end{bmatrix} 
\end{align}
%
where each submatrix $\mathbf{Q}_{i,j}'$ is the $T \times T$ sample covariance
matrix for replicates $i$ and $j$.

@@

\begin{equation}
	Q_{r,t,s} =  
		\begin{dcases}
			Q_{r,t,s}' - b_{r,t} - b_{r,t+1}, & \text{if  } t = s \\
      Q_{r,t,s}' + b_{r,\max(t,s)}, & \text{if  } |t - s| = 1 \\
		\end{dcases}
\end{equation}

where $b_{r,t}$ is element in row $r$ and column $t$ of $\mathbf{B}$.
Additionally, in some study designs, a single timepoint is shared for the
initial generation across replicates. In this case, the sampling noise is
shared between

@@

We have investigated the effectiveness of our correction on real data by
exploiting the relationship between sampling depth and the magnitude of the
variance and covariance biases, and comparing the observed variances and
covariances before and after correction. We plot the variance and covariance
(between adjacent timepoints) before and after the bias correction against the
average sample depth in 100kb genomic windows in Figure
\ref{suppfig:barghi-correction}. Overall, we find the correction strongly

>>>>

Since the seminal theoretic model of \textcite{Maynard_Smith1974-lc}
demonstrating that diversity is reduced at linked neutral variation as an
advantageous polymorphism sweeps to fixation, over four decades of theoretical
and empirical research has enriched our understanding of linked selection. This
work has led to the discovery of selected genes from genomic signals of
hitchhiking \parencite{Nair2003-tw,Voight2006-rn}, found genome-wide signals of
linked selection
\parencite{Aguade1989-jx,Begun1992-ey,Cutter2010-gi,Andersen2012-bj,Cutter2003-tl}
leading to the development of recurrent hitchhiking \parencite{Stephan1992-jc}
and background selection \parencite{Charlesworth1993-gb} models that attempt
explain this pattern. Since, much attention has focused on quantifying the
impact of background selection \parencite{McVicker2009-ax} and developing
theoretic machinery \parencite{Coop2012-cd} to differentiate the regional
reductions in diversity into those caused by full and partial sweeps, and those
caused by background selection \parencite{Elyashiv2016-vt}. However, as other
theoretic models of soft sweeps from standing variation emerged
\parencite{Hermisson2005-hs,Pennings2006-lj}, attention shifted towards
detecting other modes of linked selection \parencite{Pritchard2010-tk}, a more
difficult endeavor \parencite{Przeworski2005-bg} met with new coalescent models
\parencite{Berg2015-xj} and machine learning methods
\parencite{Schrider2017-yx}. An undervalued theoretic approach to understanding
the genome-wide effects of selection on standing variation, e.g. selection on
an infinitesimal polygenic trait, stems from an early quantitative genetic
model of linked selection \parencite{Robertson1961-ho} and its later
developments \parencite{Santiago1995-hx,Santiago1998-bs}. Implicit in these
models is that autocovariance between allele frequency change is created when
there is heritable fitness variation in the population, a signal that can be
readily detected from temporal genomic data \parencite{Buffalo2019-io}. Here,
we provide the first empirical evidence of linked selection acting on standing
variation over very short timescales.

We reanalyzed three published temporal genomic datasets and detected the
genome-wide signature of linked selection acting over tens of generations.
Furthermore, we find the dynamics of temporal autocovariance are consistent
with theory; the temporal covariance in allele frequencies decays \vb{as both
the fitness variation associated with particular haplotypes and the linkage
disequilibria between selected and neutral loci decay through time}. In
reanalyzing one study \parencite{Barghi2019-qy}, we find that after sixty
generations, greater than 20\% of the variation in allele frequency change is
directly due to the action linked selection. Capitalizing on replicated
evolve-and-resequence study designs, we characterized the extent to which
convergent selection pressures lead to parallel changes in allele frequency
across replicate populations, finding moderate correlations in allele frequency
changes.

@@

It is worth building some intuition why temporal covariance allows us to detect
such faint signals of polygenic linked selection from temporal genomic data.
Each variant is subject to both variance in allele frequency due to drift and
two levels of sampling noise (as individuals are sampled from the population
and reads are sampled from amplified DNA), which at any locus swamps the
temporal covariance signal and creates spurious covariances. However, these
spurious covariances do not share a common sign across timepoints whereas the
covariances created by linked selection do; consequently, averaging across the
entire genome, the temporal signal exceeds sampling noise.

One limitation of our current work here is that none of the studies we
reanalyzed had linkage disequilibria data for the evolved populations. Our
theory of temporal autocovariance tells us that the temporal autocovariances a
neutral site experiences is determined by the product of the expected linkage
disequilibrium and additive fitness variation that persists through the
generations \parencite{Buffalo2019-io}. This leads to a clear prediction:
regions of higher linkage disequilibrium and lower recombination should have
greater temporal autocovariance than regions with lower LD and higher
recombination. Unfortunately, we lack a high-resolution recombination map for
\emph{D. simulans} and while there are LD data for \emph{D. simulans}
\parencite{Signor2018-wg} we did not find a relationship between temporal
covariance and LD.

% \graham{TODO Is important point. We'll likely need plots for supp}

@@

Thus far, the most comprehensive studies implicating linked selection in
affecting genome-wide diversity have been in \emph{Drosophila}
\parencite{Begun1992-ey,Elyashiv2016-vt,Sattath2011-dr} and
\emph{Caenorhabditis} \parencite{Cutter2003-tl,Cutter2003-tl,Andersen2012-bj},
whereas the evidence for linked selection in plant and yeast species is weak
\parencite{Cutter2013-ba}. Both \emph{Drosophila} and \emph{Caenorhabditi} have
short genetic maps (or in the case of selfing \emph{C.  elegans}, small
effective recombination rates), and large effect population sizes, two factors
predicted to increase the degree to which hitchhiking impacts genome-wide
diversity \parencite{Barton2000-zg}. However, our results here show that even
with small effective population sizes (300, 450, and 45 for the
\textcite{Kelly2019-dc}, and \textcite{Castro2019-uk} studies respectively)
around and, in the case of mice, moderately-sized genetic maps (around 14
Morgans; \cite{Cox2009-hf}), we still see a fairly strong effect of linked
selection. This suggests that while these factors may govern the dynamics of
classic hitchhiking events, they could perhaps play less of a role in polygenic
linked selection. However, with only three cases here, further studies on
different taxa are needed.

@@

Following \textcite{Waples1989-sj}, we have that that the variance in allele
frequency change at a locus in the initial generation, which is entirely due to
the binomial sampling process, is $\var(p_0) = \nicefrac{p_0(1-p_0)}{d_0}$
where $d_0$ is the number of binomial draws (e.g. read depth). At a later
timepoint, the variance in allele frequency is a result of both the binomial
sampling process at time $t$ and the evolutionary process. Using the law of
total variation we can partition the variation from each process,

\begin{align}
  \var(\widetilde{p_t}) &= \E(\var(\widetilde{p_t} | p_t)) + \var(\E(\widetilde{p_t}|p_t)) \\
                        &= \underbrace{\frac{p_t(1-p_t)}{d_t}}_\text{generation $t$ sampling noise} + \underbrace{\var(p_t)}_\text{variance due to evolutionary process}.
  %\frac{\var(\widetilde{p_t})}{p_0(1-p_0)} &= \frac{p_t(1-p_t)}{p_0(1-p_0) d_t} + 1 - \left( 1-\frac{1}{2N}\right)^t.
  %\frac{\var(\widetilde{p_t})}{p_0(1-p_0)} &= \frac{p_t(1-p_t)}{p_0(1-p_0) d_t} + \frac{t}{2N} + O\left((2 N)^{-2}\right).
\end{align}

@@

\begin{align}
  \sum_{i=0}^{t-1} \var(\Delta_{_D} p_i) = \sum_{i=0}^{t-1} \frac{p_i(1-p_i)}{2N}
\end{align}
%
replacing the heterozygosity in generation $i$ with its expectation, we have

\begin{align}
  \sum_{i=0}^{t-1} \var(\Delta_{_D} p_i) &= p_0(1-p_0) \sum_{i=0}^{t-1} \frac{1}{2N} \left(1-\frac{1}{2N}\right)^i \\
                                         &= p_0(1-p_0) \left[1 - \left(1-\frac{1}{2N}\right)^t \right]
\end{align}
%
which is the usual variance in allele frequency change due to drift.  Then, the
total allele frequency change from generations $0$ to $t$ is
$\var(\widetilde{p}_t - \widetilde{p}_0) = \var(\widetilde{p}_t) +
\var(\widetilde{p}_0) - 2 \cov(\widetilde{p}_t, \widetilde{p}_0)$, where the
covariance depends on the nature of the sampling plan (see \cite{Nei1981-oy,
Waples1989-sj}). In the case where there is heritable variation for fitness,
and using the fact that $\cov(\widetilde{p}_t, \widetilde{p}_0) =
\nicefrac{p_0(1-p_0)}{2N}$ for Plan I sampling procedures
\parencite{Waples1989-sj}, we write,

\begin{align}
  \var(\widetilde{p}_t - \widetilde{p}_0) &= \var(\widetilde{p}_t) + \var(\widetilde{p}_0) - 2 C \cov(\widetilde{p}_t, \widetilde{p}_0) \\
                                          &= \frac{p_t(1-p_t)}{d_t}  + \frac{p_0(1-p_0)}{d_0} + p_0(1-p_0) \left[1 - \left(1-\frac{1}{2N}\right)^t \right] + \\ & \;\;\;\;\;\;
                                               \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i)  + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j) - \frac{C p_0(1-p_0)}{2N} \\
  \frac{\var(\widetilde{p}_t - \widetilde{p}_0)}{p_0(1-p_0)} &= 1 + \frac{p_t(1-p_t)}{p_0(1-p_0)d_t}  + \frac{1}{d_0} - \left(1-\frac{1}{2N}\right)^t + \\ & \;\;\;\;\;\;
  \sum_{i=0}^{t-1} \frac{\var(\Delta_{_H} p_i)}{p_0(1-p_0)}  + \sum_{0 \le i < j}^{t-1} \frac{\cov(\Delta p_i, \Delta p_j)}{p_0(1-p_0)} - \frac{C}{N}
\end{align}
%
where $C = 1$ if Plan I is used, and $C=0$ if Plan II is used (see
\cite{Waples1989-sj}, p. 380 and Figure 1 for a description of these sampling
procedures; throughout the paper we use sampling Plan II). We move terms
creating a bias-corrected estimator for the population variance in allele
frequency change, and replace all population heterozygosity terms with the
unbiased sample estimators, e.g. $\frac{d_t}{d_t-1} \widetilde{p}_t (1-
\widetilde{p}_t)$,

@@

\begin{align}
  \var(\widetilde{p}_t - \widetilde{p}_0) &= 
  p_t(1-p_t) \left(\frac{1}{n_t} + \frac{1}{d_t} - \frac{1}{n_t d_t} \right)  
  + p_0(1-p_0) \left( \frac{1}{n_0} + \frac{1}{d_0} - \frac{1}{n_0 d_0}\right)  \\ & \;\;\;\;\;\;
  - \frac{C p_0(1-p_0)}{N} + p_0(1-p_0) \left[1 - \left(1-\frac{1}{2N}\right)^t \right]+ \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i)  \\ & \;\;\;\;\;\; + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j) 
\end{align}
%
Through the law of total expectation (see \cite{Kolaczkowski2011-ee}
Supplementary File 1 for a sample proof), one can find that an unbiased
estimator of the half the heterozygosity is

\begin{align}
  \frac{n_t d_t}{(n_t-1) (d_t-1)} \widetilde{p_t}(1-\widetilde{p_t}).
\end{align}
%
Replacing this unbiased estimator for half of the heterozygosity into our
expression above, the total sample variance is

\begin{align}
  \var(\widetilde{p}_t - \widetilde{p}_0) &= 
  \frac{n_t d_t \widetilde{p}_t(1-\widetilde{p}_t)}{(n_t-1)(d_t-1)} \left(\frac{1}{n_t} + \frac{1}{d_t} - \frac{1}{n_t d_t} \right) + 
 \frac{n_0 d_0 \widetilde{p}_0(1-\widetilde{p}_0)}{(n_0-1)(d_0-1)} \left( \frac{1}{n_0} + \frac{1}{d_0} - \frac{1}{n_0 d_0}\right) + \\ & \nonumber\;\;\;\;\;\;
 \frac{n_0 d_0 \widetilde{p}_0(1-\widetilde{p}_0)}{(n_0-1)(d_0-1)}   \left[1 - \left(1-\frac{1}{2N}\right)^t \right]  - \frac{C}{N}  \frac{n_0 d_0 \widetilde{p}_0(1-\widetilde{p}_0)}{(n_0-1)(d_0-1)} + \\ \nonumber & \;\;\;\;\;\; \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i)  + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j).  \\
                                                                                                                      % &= \widetilde{p}_t(1-\widetilde{p}_t)\frac{d_t + n_t - 1}{(n_t-1)(d_t-1)} + 
 % \widetilde{p}_0(1-\widetilde{p}_0)\frac{d_0 + n_0 - 1}{(n_0-1)(d_0-1)} + \\ & \nonumber\;\;\;\;\;\;
 % \widetilde{p}_0(1-\widetilde{p}_0) \frac{n_0 d_0}{(n_0-1)(d_0-1)}  \left[1 - \left(1-\frac{1}{2N}\right)^t \right] - \frac{C}{N} \widetilde{p}_0(1-\widetilde{p}_0)\frac{n_0 d_0}{(n_0-1)(d_0-1)} 
 % \\ \nonumber & \;\;\;\;\;\; + \sum_{i=0}^{t-1} \var(\Delta_{_H} p_i)  + \sum_{0 \le i < j}^{t-1} \cov(\Delta p_i, \Delta p_j). 
\end{align}
%
As with equation \eqref{supp:eqn-depth-only-correction}, we can rearrange this
to get a biased-correct estimate of the variance in allele frequency change
between adjacent generations, $\var(\Delta p_t)$.

@@

We also need to apply a bias correction to the temporal covariances (and
possibly the replicate covariances if the initial sample frequencies are all
shared). The basic issue is that $\cov(\Delta \widetilde{p}_t, \Delta
\widetilde{p}_{t+1}) = \cov(\widetilde{p}_{t+1} - \widetilde{p}_t,
\widetilde{p}_{t+2} - \widetilde{p}_{t+1})$, and thus shares the sampling noise
of timepoint $t+1$. Thus acts to bias the covariance by subtracting off the
noise variance term of $\var(\widetilde{p}_{t+1})$, so we add the expectation
of this bias, derived above, back in. We discuss this in more detail below in
deriving the bias correction for the temporal-replicate variance covariance
matrix.

\subsection{Temporal-Replicate Covariance Matrix Correction}
\label{supp:matrix-correction}

In practice, we simultaneously estimate the temporal and replicate covariance
matrices for each replicate, which we call the temporal-replicate covariance
matrix. This needs a bias correction; we extend the bias corrections for single
locus variance and covariance described in Supplementary Material Sections
\ref{supp:depth-var-corr}, \ref{supp:ind-depth-var-corr}, and
\ref{supp:cov-corr} to multiple sampled loci and the temporal-replicate
covariance matrix here. With frequency data collected at $T+1$ timepoints
across $R$ replicate populations at $L$ loci, we have multidimensional arrays
$\mathbf{F}$ of allele frequencies, $\mathbf{D}$ of sequencing depths, and
$\mathbf{N}$ of the number of individuals sequenced, each of dimension $R
\times (T+1) \times L$.  We calculate the array $\mathbf{\Delta F}$ which
contains the allele frequency changes between adjacent generations, and has
dimension $R \times T \times L$.  The operation
$\flt(\mathbf{\Delta}\mathbf{F})$ flattens this array to a $(R \cdot T) \times
L$ matrix, such that rows are grouped by replicate, e.g. for timepoint $t$,
replicate $r$, and locus $l$ such that for allele frequencies $p_{t, r, l}$,
the frequency change entries are

@@

\begin{align}
    \flt(\mathbf{\Delta F}) &=
                    &\begin{bmatrix} 
    \Delta p_{1, 0, 0} & \Delta p_{2, 0, 0} & \ldots & \Delta p_{1, 1, 0} & \Delta p_{2, 1, 0} & \ldots & \Delta p_{T, R, 0}  \\
    \Delta p_{1, 0, 1} & \Delta p_{2, 0, 1} & \ldots & \Delta p_{1, 1, 1} & \Delta p_{2, 1, 1} & \ldots & \Delta p_{T, R, 1}  \\
    \vdots & \vdots & \ddots & \vdots & \vdots & \ddots & \vdots  \\
    \Delta p_{1, 0, L} & \Delta p_{2, 0, L} & \ldots & \Delta p_{1, 1, L} & \Delta p_{2, 1, L} & \ldots & \Delta p_{T, R, L}  \\
  \end{bmatrix} 
\end{align}
%
where each $\Delta p_{t, r, l} = p_{t+1, r, l} - p_{t, r, l}$. Then, the sample
temporal-replicate covariance matrix $\mathbf{Q}'$ calculated on
$\flt(\mathbf{\Delta F})$ is a $(R \cdot T) \times (R \cdot T)$ matrix, with
the $R$ temporal-covariance block submatrices along the diagonal, and the
$R(R-1)$ replicate-covariance submatrices matrices in the upper and lower
triangles of the matrix,

\begin{align}
	\mathbf{Q}' &= 
  \begin{bmatrix} 
		\mathbf{Q}_{1,1}' & \mathbf{Q}_{1, 2}' & \ldots & \mathbf{Q}_{1, R}' \\ 
		\mathbf{Q}_{2,1}' & \mathbf{Q}_{2, 2}' & \ldots & \mathbf{Q}_{2, R}' \\ 
		\vdots & \vdots & \ddots & \vdots \\
		\mathbf{Q}_{R,1}' & \mathbf{Q}_{R, 2}' & \ldots & \mathbf{Q}_{R, R}' \\ 
  \end{bmatrix} 
\end{align}
%
where each submatrix $\mathbf{Q}_{i,j}'$ ($i \ne j$) is the $T \times T$ sample
replicate covariance matrix for replicates $i$ and $j$, and the submatrices
along the diagonal $\mathbf{Q}_{r,r}'$ are the temporal covariance matrices for
replicate $r$.

@@

\begin{equation}
	Q_{r,t,s} =  
		\begin{dcases}
			Q_{r,t,s}' - b_{r,t} - b_{r,t+1}, & \text{if  } t = s \\
      Q_{r,t,s}' + b_{r,\max(t,s)}, & \text{if  } |t - s| = 1 \\
		\end{dcases}
\end{equation}
%
where $b_{r,t}$ is element in row $r$ and column $t$ of $\mathbf{B}$.

%Additionally, in some study designs, a single timepoint is shared for the
%initial generation across replicates. In this case, the sampling noise is
%shared between 
% TODO

@@

We have investigated the effectiveness of our correction on real data by
exploiting the relationship between sampling depth and the magnitude of the
variance and covariance biases, and comparing the observed variances and
covariances before and after correction. We plot the variance and covariance
(between adjacent timepoints) before and after the bias correction against the
average sample depth in 100kb genomic windows in Figure
\ref{suppfig:barghi-correction}. Overall, we find the biased-correction
procedure removes the relationship between variance and covariance and depth, indicating it is working adequately.
